\mythesistitle

### 2011-11-22 src/synopsis.tex
a whole bunch of changes i cannot rememeber now. deal with it.

@@

% \tnote{testing 1 2 3...}
% \tnote[bogus1]{testing 1 2 3...}
% \tnote[bogus2]{testing 1 2 3...}
% \tnote[bogus3]{testing 1 2 3...}
% \tnote[bogus4]{testing 1 2 3...}
% \tnote[bogus5]{testing 1 2 3...}
% \tnote[bogus6]{testing 1 2 3...}
% \tnote[bogus7]{testing 1 2 3...}
% \tnote[bogus8]{testing 1 2 3...}
% \tnote[TH]{testing 1 2 3...}

@@

A group of students, \Pa, \Pi, \Sn, \Wo, \Vi, \Li, \Ch, \Sa, \Fr,
  \Sc\ and \Lu\ enroll at the
% {\bf Wo}odstock, {\bf Vi}olet, {\bf Li}nus, {\bf Ch}arlie, {\bf
%   Sa}lly, {\bf Re}run, {\bf Sc}hr{\"o}eder} and {\em {\bf Lu}cy}
% enroll at the
{\WSI} for a liberal arts programme.  As part
of their semester thesis, they pick a body of work to study and form
the namesake study groups, {\LLL}, {\GGG}, {\BBB} and {\TTT}. A student
will be in at least one study group and may be in more than one. For
instance, as will be seen later, {\Fr} studies both {\LLL} and {\TTT}
while \Wo\ studies only \BBB.

@@

\begin{figure}[htbp]
 \centering
  \includegraphics[scale=0.3]{../img/1_infinite_loop.png}
  \caption{{\residenceblock} street map}
  \label{fig:streetmap}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.3]{../img/2_infinite_loop_BTWF.png}
  \caption{{\residenceblock} street map with study group routes
    allocated. Routes are color coded as follows: red for
    \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
    group, orange for \textcolor{YellowOrange}{$\xBBB$} group, green for
    \textcolor{green}{$\xTTT$} group}
  \label{fig:streetmappath}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../img/3_infinite_loop.png}
  \caption{Individual allocation of apartments to students in
    {\residenceblock} that meets the requirements stated before.}
  \label{fig:streetmappathpeople}
\end{figure}

@@

The interval assignment problem to a set system is equivalent to the
consecutive ones problem (COP) in binary matrices\cite{wlh02,
  nsnrs09}.  The COP problem is to rearrange rows (columns) of a
binary matrix in such a way that every column (row) has its $1$s occur
consecutively. If this is possible the matrix is said to have the COP.
Consecutive ones property (COP) of binary matrices is a well
researched combinatorial problem and has several positive results on
tests for it and computing the COP permutation (i.e. the course
schedule in the above illustration) which will be surveyed later in
this document. Hence we are interested in extensions of COP, more
specifically, the extension of interval assignment problem to tree
path assignment problem (which is illustrated by the study group
accommodation problem).

@@

%The course scheduling problem illustated earlier, as a special case of

It is quite obvious now why this version of the
problem (where the ``target graph'' is a path and not any
tree\tnote[TH]{Allowing any tree in this example could be seen as a
  scenario where there are parallel classes. A node falling in the
  path between two other nodes would mean that the corresponding is
  scheduled between the other two.}) is called an interval assignment
problem.

The interval assignment problem to a set system is equivalent to the
consecutive ones problem (COP) in binary matrices\cite{wlh02,
  nsnrs09}.  The COP problem is to rearrange rows (columns) of a
binary matrix in such a way that every column (row) has its $1$s occur
consecutively. If this is possible the matrix is said to have the COP.
Consecutive ones property (COP) of binary matrices is a well
researched combinatorial problem and has several positive results on
tests for it and computing the COP permutation (i.e. the course
schedule in the above illustration) which will be surveyed later in
this document. Hence we are interested in extensions of COP, more
specifically, the extension of interval assignment problem to tree
path assignment problem (which is illustrated by the study group
accommodation problem).

As described earlier, 
a matrix with COP is one whose rows (columns)
can be rearranged so that the 1s in every column (row) are in
consecutive rows (columns).

\def \un {\bf 1}
\begin{figure}[htbp]
  \centering
  % \includegraphics[scale=0.4]{../img/0_cop_nocop.png}
  \begin{tabular}[h]{l|lcccl}
    $M_1$: & $M_1'$: &&&& $M_2$:\\
    &&&&&\\
    \begin{tabular}[h]{llll}
      $c_1$ & $c_2$ &$c_3$ &$c_4$\\
      &&&\\
      \un & 0   & \un & 0\\
      0   & \un & 0   & \un \\
      \un & 0   & 0   & \un
    \end{tabular}
    &
    \begin{tabular}[h]{llll}
      $c_3$ &$c_1$ &$c_4$& $c_2$\\
      &&&\\
      \un & \un & 0 & 0\\
      0 & 0 & \un & \un \\
      0 & \un & \un & 0
    \end{tabular}
    &&&&
    \begin{tabular}[h]{llll}
      $d_1$ & $d_2$ &$d_3$ &$d_4$\\
      % &&&\\
      &&&\\
      \un & \un & 0 & 0\\
      0 & \un & \un & 0 \\
      0 & \un & 0 & \un 
    \end{tabular}
  \end{tabular}
  \caption{Matrices with and without COP. $M_1$ has COP because by
    permuting its columns, $c_1$-$c_4$, one can obtain $M_1'$ where
    the {\un}s in each row are consecutive. $M_2$, however, does not
    have COP since no permutation of its columns, $d_1$-$d_4$, will
    arrange {\un}s in each row consecutively \cite{d08phd}.}
  \label{fig:cop-matrix}
\end{figure}

COP in binary matrices has several practical applications in diverse
fields including scheduling \cite{hl06}, information retrieval
\cite{k77} and computational biology \cite{abh98}.  Further, it is a
tool in graph theory \cite{mcg04} for interval graph recognition,
characterization of Hamiltonian graphs, planarity testing \cite{bl76}
and in integer linear programming \cite{ht02,hl06}.

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  Since then several algorithms have been
invented; PQ trees \cite{bl76} gave the first linear time
algorithm. Variations of PQ trees \cite{mm96,wlh01,wlh02,mcc04} and
ICPIA \cite{nsnrs09} further improve either the time complexity and/or
the complexity of implementation.

\cite{bl76} describes a linear time construction of a data structure
called PQ trees. This paper draws on the close relationship of COP
matrices to interval graphs. A PQ tree of a matrix is one that stores
all row (column) permutations of the matrix that give the COP orders
(there could be multiple) of the matrix. This is constructed using an
elaborate linear time procedure. Since this result, several
improvements have been made to the procedural complexity of the
algorithm itself and also derivative data structures that are easier
to construct. PQR trees is one such data structure \cite{mm96},
\cite{mpt98}\tnote[TH]{also, mpt98 $==$ elaborate version of mm96}.
\cite{tm05} describes an improved version of PQR
trees. \tnote{improv in terms of what?}\cite{wlh02} describes the
simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote[TH]{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great
degree. \tnote{get this
  figured out! - because it reduces the search space for the solution
  from set of all permutations of rows to set of all}

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property and
if not it exited stating negative, this algorithm gives an evidence by
way of a certificate of
matrix being devoid of COP thereby enabling a user to check the
algorithm's result. This is significant from an implementation
perspective because automated program verification is hard and manual
verification is more viable. Hence having a certificate
reinforces its credibility. Note that when the matrix has COP the COP
order is the certificate.  It may also be noted that the internal
machinery of this algorithm \annote {is related to the weighted
  betweenness problem addressed } {in what way??} in \cite{co98}.
\tnote{expand on the COP order graph creation and it having to be
  bipartite for M to have COP. and thus an odd cycle being an evidence
  of no COP.}
\tnote[TH]{ where should this go?: cite|jlm97 (application of PQ trees
  in graphics).\\ add helly's theorem citation
  19XXdgk-Hellystheorem-Danzer-Gruenbaum-Klee}

The literature also has results on parallel COT algorithms --
\cite{as95} \cite{bs03} \cite{ly91}\tnote{add english to fit context}

\subsection{Optimization problems in COP}
So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP. Nevertheless, the COP is
not arbirary and there are several interesting problems when a matrix
does not have COP but is ``close'' to having COP or can be altered to
have COP.

-- sect 4.1 in \cite{d08phd} has many results surveyed. hardness
results, approx. results. results are usually for a class of matrices
$(a,b)$ where number columns and rows are restriced to $a$ and $b$
\tnote{check if it is the other way around}.\\
-- problem of flipping at most $k$ entries of $M$ to make it attain
COP. this is NP complete \cite{b75-phd}

The first step in this direction is to analyze matrices without
COP. \cite{at72} showed that a matrix that does not have COP have
certain forbidden substructures that prevent it from having
COP. Tucker classified these substructures into five classes of
submatrices. This result is presented in the context of AT-free
bipartite graphs \tnote[TH]{ check this up. give details.} which is
equivalent to matrices with COP. A binary matrix is converted into a
bipartite graph as follows. The set of rows and the set of columns
form the two partitions of the graph. Each row node is adjacent to
those nodes that represent the columns that have ones in the
corresponding row. \cite{at72} proves that this bipartite graph has no
asteroidal triple if and only if the matrix has COP and goes on to
identify the forbidden substructures for these bipartite graphs. The
matrices corresponding to these substructures are the forbidden
submatrices. Once a matrix has been detected to not have COP (using
any of the COT algorithms mentioned earlier), it is naturally of
interest to find out the smallest (in terms of number of rows, columns
and number of entries that are ones) forbidden
substructure. \cite{d08phd} discusses an exact algorithm with time
complexity \annote{parameterized by the number of ones in the
  matrix}{check this.}.

@@

How the problem of COT is related to the constraint statisfaction
problem of interval assignment to sets in a system can be seen as
follows. Every column of the binary matrix can be converted into a set
of integers which are the indices of rows with $1$s in that column. It
is apparent that if the matrix has COP, then constructing such sets
after applying the COP permutation to the matrix will result in all
sets with consecutive integers. In other words, after application of
COP reordering, the sets are intervals. Indeed the problem now becomes
finding interval assignments to a given set system such that there
exists a permutation of the universe (set of row indices) which
converts each set to its assigned interval. The result in
\cite{nsnrs09} characterizes interval assignments to the sets which
can be obtained from a single permutation of the rows.  They show that
for each set, the cardinality of the interval assigned to it must be
same as the cardinality of the set, and the intersection cardinality
of any two sets must be same as the intersection cardinality of the
corresponding intervals.  While this is obviously a necessary
condition, this result shows this is also sufficient.  \cite{nsnrs09}
calls such an interval assignment an Intersection Cardinality
Preserving Interval Assignment (ICPIA).  This paper generalizes the
idea from \cite{wlh02} of decomposing a given binary matrix into prime
matrices for COT and describes an algorithm to test if an ICPIA exists
for a given set system.

>>>>

\def \un {\bf 1}

@@

\tnote{testing 1 2 3...}
\tnote[bogus1]{testing 1 2 3...}
\tnote[bogus2]{testing 1 2 3...}
\tnote[GTC]{testing 1 2 3...}
\tnote[bogus4]{testing 1 2 3...}
\tnote[TH]{testing 1 2 3...}

@@

A group of students, \Pa, \Pi, \Sn, \Wo, \Vi, \Li, \Ch, \Sa, \Fr,
  \Sc\ and \Lu\ enroll at the
% {\bf Wo}odstock, {\bf Vi}olet, {\bf Li}nus, {\bf Ch}arlie, {\bf
%   Sa}lly, {\bf Re}run, {\bf Sc}hr{\"o}eder} and {\em {\bf Lu}cy}
% enroll at the
{\WSI} for a liberal arts programme.  As part
of their semester thesis, they pick a body of work to study and form
the namesake study groups, {\LLL}, {\GGG}, {\BBB} and
{\TTT}. \tnote[TH]{put bib entries for these works!} A student
will be in at least one study group and may be in more than one. For
instance, as will be seen later, {\Fr} studies both {\LLL} and {\TTT}
while \Wo\ studies only \BBB.

@@

\begin{figure}[t]%[htbp]
 \centering
  \includegraphics[scale=0.3]{../img/1_infinite_loop.png}
  \caption{{\residenceblock} street map}
  \label{fig:streetmap}
\end{figure}

\begin{figure}[t]%[htbp]
  \centering
  \includegraphics[scale=0.3]{../img/2_infinite_loop_BTWF.png}
  \caption{{\residenceblock} street map with study group routes
    allocated. Routes are color coded as follows: red for
    \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
    group, orange for \textcolor{YellowOrange}{$\xBBB$} group, green for
    \textcolor{green}{$\xTTT$} group}
  \label{fig:streetmappath}
\end{figure}

\begin{figure}[t] % [h]
  \centering
  \includegraphics[scale=0.5]{../img/3_infinite_loop.png}
  \caption{Individual allocation of apartments to students in
    {\residenceblock} that meets the requirements stated before.}
  \label{fig:streetmappathpeople}
\end{figure}

@@

The interval assignment problem to a set system is equivalent to the
consecutive ones problem (COP) in binary matrices\cite{wlh02,
  nsnrs09}.  The COP problem is to rearrange rows (columns) of a
binary matrix in such a way that every column (row) has its {\un}s occur
consecutively. If this is possible the matrix is said to have the COP.
Consecutive ones property (COP) of binary matrices is a well
researched combinatorial problem and has several positive results on
tests for it and computing the COP permutation (i.e. the course
schedule in the above illustration) which will be surveyed later in
this document. Hence we are interested in extensions of COP, more
specifically, the extension of interval assignment problem to tree
path assignment problem (which is illustrated by the study group
accommodation problem).

@@

As seen earlier, the interval assignment problem (illustrated as the
course scheduling problem in Section~\ref{sec:problem}), is a special
case of the problem we address in this thesis, namely the tree path
labeling problem (illustrated as the study group accommodation
problem). Interval assignment and COP are equivalent
problems. In this section we will see some of the results that
exists in the literature today towards solving the COP problem and
optimization problems surrounding it.

% The interval assignment problem to a set system is equivalent to the
% consecutive ones problem (COP) in binary matrices\cite{wlh02,
%   nsnrs09}.  The COP problem is to rearrange rows (columns) of a
% binary matrix in such a way that every column (row) has its {\un}s occur
% consecutively. If this is possible the matrix is said to have the COP.
% Consecutive ones property (COP) of binary matrices is a well
% researched combinatorial problem and has several positive results on
% tests for it and computing the COP permutation (i.e. the course
% schedule in the above illustration) which will be surveyed later in
% this document. Hence we are interested in extensions of COP, more
% specifically, the extension of interval assignment problem to tree
% path assignment problem (which is illustrated by the study group
% accommodation problem).

\begin{figure}[t] %[htbp]
  \centering
  % \includegraphics[scale=0.4]{../img/0_cop_nocop.png}
  \begin{tabular}[h]{l|lcccl}
    $M_1$: & $M_1'$: &&&& $M_2$:\\
    &&&&&\\
    \begin{tabular}[h]{llll}
      $c_1$ & $c_2$ &$c_3$ &$c_4$\\
      &&&\\
      \un & 0   & \un & 0\\
      0   & \un & 0   & \un \\
      \un & 0   & 0   & \un
    \end{tabular}
    &
    \begin{tabular}[h]{llll}
      $c_3$ &$c_1$ &$c_4$& $c_2$\\
      &&&\\
      \un & \un & 0 & 0\\
      0 & 0 & \un & \un \\
      0 & \un & \un & 0
    \end{tabular}
    &&&&
    \begin{tabular}[h]{llll}
      $d_1$ & $d_2$ &$d_3$ &$d_4$\\
      % &&&\\
      &&&\\
      \un & \un & 0 & 0\\
      0 & \un & \un & 0 \\
      0 & \un & 0 & \un 
    \end{tabular}
  \end{tabular}
  \caption{Matrices with and without COP. $M_1$ has COP because by
    permuting its columns, $c_1$-$c_4$, one can obtain $M_1'$ where
    the {\un}s in each row are consecutive. $M_2$, however, does not
    have COP since no permutation of its columns, $d_1$-$d_4$, will
    arrange {\un}s in each row consecutively \cite{d08phd}.}
  \label{fig:cop-matrix}
\end{figure}

Recall that a matrix with COP is one whose rows (columns)
can be rearranged so that the {\un}s in every column (row) are in
consecutive rows (columns). Figure~\ref{fig:cop-matrix} shows examples
of this property.
COP in binary matrices has several practical applications in diverse
fields including scheduling \cite{hl06}, information retrieval
\cite{k77} and computational biology \cite{abh98}.  Further, it is a
tool in graph theory \cite{mcg04} for interval graph recognition,
characterization of Hamiltonian graphs, planarity testing \cite{bl76}
and in integer linear programming \cite{ht02,hl06}.

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  A landmark result came a few years later
when \cite{at72} discovered the families of forbidden submatrices that
prevent a matrix from having COP and most, if not all, results that
came later were based on this discovery which connected COP in binary
matrices to convex bipartite graphs. In fact, the forbidden
submatrices came as a corollary to the discovery that convex bipartite
graphs are AT-free in \cite{at72}\tnote{check}. The first linear time
algorithm for COP testing (COT) was invented by \cite{bl76} using a
data structure called PQ trees.  Since then several COT algorithms
have been invented -- some of which involved variations of PQ trees
\cite{mm96,wlh01,mcc04}, some involved set theory and ICPIA
\cite{wlh02,nsnrs09}, parallel COT algorithms\cite{as95,bs03,ly91} and
certifying algorithms\cite{mcc04}. % These algorithms
% improve over PQ trees on the time complexity and/or the complexity
% of implementation.

The construction of PQ trees in \cite{bl76} draws on the
close relationship of COP matrices to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple) of the matrix. This
is constructed using an elaborate linear time procedure and is also
a test for planarity\tnote{check check check. both interval
  graph and planarity in this paper?}. % Since this
% result, several improvements have been made to the procedural
% complexity of the algorithm itself and also derivative data structures
% that are easier to construct.
PQR trees is a generalized data structure based on PQ trees
\cite{mm96,mpt98}.  \cite{tm05} describes an improved algorithm to
build PQR trees. \tnote[GTC]{improv in terms of what?}\cite{wlh02}
describes the simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote[TH]{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great
degree. \tnote[GTC]{it reduces the solution search space. fill in the blanks.}

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed\tnote[GTC]{in what way??} in \cite{co98}.  \tnote[GTC]{expand
  on the COP order graph creation and it having to be bipartite for M
  to have COP. and thus an odd cycle being an evidence of no COP.}
\tnote[TH]{ where should this go?: (1) cite|jlm97 (application of PQ trees
  in graphics). (2) helly's theorem citation
  19XXdgk-Hellystheorem-Danzer-Gruenbaum-Klee}

\subsection{Optimization problems in COP}
\label{sec:optcop}
So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP, primarily due to the
noisy nature of data available. At the same time, COP is not arbitrary
and is a desirable property in practical data representation
\cite{co98,jkckv04,k77}. In this context, there are several
interesting problems when a matrix does not have COP but is ``close''
to having COP or is allowed to be altered to have COP. These are the
optimization problems related to a matrix which does not have
COP. Some of the significant problems are surveyed in this section.

MOVE TO THESIS?
-- sect 4.1 in \cite{d08phd} has many results surveyed. hardness
results, approx. results. results are usually for a class of matrices
$(a,b)$ where number columns and rows are restriced to $a$ and $b$
\tnote{check if it is the other way around}.\\
-- problem of flipping at most $k$ entries of $M$ to make it attain
COP. this is NP complete \cite{b75-phd}
END MOVE TO THESIS?

\tnote[TH]{cite:lb62 - interval graphs are at-free chordal}

\cite{at72} showed that a matrix that does not have COP have
certain substructures that prevent it from having
COP. Tucker classified these forbidden substructures into five classes of
submatrices. This result is presented in the context of convex
bipartite graphs which \cite{at72} proved to be
AT-free\tnote[TH]{ check this up. give details. - doms'}. Convex bipartite
graph have vertex incidence adjacency matrices is
equivalent to matrices with COP. A binary matrix is converted into a
bipartite graph as follows. The set of rows and the set of columns
form the two partitions of the graph. Each row node is adjacent to
those nodes that represent the columns that have ones in the
corresponding row. \cite{at72} proves that this bipartite graph has no
asteroidal triple if and only if the matrix has COP and goes on to
identify the forbidden substructures for these bipartite graphs. The
matrices corresponding to these substructures are the forbidden
submatrices. Once a matrix has been detected to not have COP (using
any of the COT algorithms mentioned earlier), it is naturally of
interest to find out the smallest (in terms of number of rows, columns
and number of entries that are ones) forbidden
substructure. \cite{d08phd} discusses an exact algorithm with time
complexity \annote{parameterized by the number of ones in the
  matrix}{check this.}.

@@

How the problem of COT is related to the constraint statisfaction
problem of interval assignment to sets in a system can be seen as
follows. Every column of the binary matrix can be converted into a set
of integers which are the indices of rows with {\un}s in that column. It
is apparent that if the matrix has COP, then constructing such sets
after applying the COP permutation to the matrix will result in all
sets with consecutive integers. In other words, after application of
COP reordering, the sets are intervals. Indeed the problem now becomes
finding interval assignments to a given set system such that there
exists a permutation of the universe (set of row indices) which
converts each set to its assigned interval. The result in
\cite{nsnrs09} characterizes interval assignments to the sets which
can be obtained from a single permutation of the rows.  They show that
for each set, the cardinality of the interval assigned to it must be
same as the cardinality of the set, and the intersection cardinality
of any two sets must be same as the intersection cardinality of the
corresponding intervals.  While this is obviously a necessary
condition, this result shows this is also sufficient.  \cite{nsnrs09}
calls such an interval assignment an Intersection Cardinality
Preserving Interval Assignment (ICPIA).  This paper generalizes the
idea from \cite{wlh02} of decomposing a given binary matrix into prime
matrices for COT and describes an algorithm to test if an ICPIA exists
for a given set system.
